---
title: "Implement of IBaySVG"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
---

# 1. Overview

This document provides a detailed user guide for the **IBaySVG** function, which implements a Bayesian framework for identifying spatially variable (SV) genes across multiple spatial transcriptomics slices.The function supports multi-slice integration, optional covariate adjustment, and parallel computation. The code is available on https://github.com/zhoumeng123456/IBaySVG.


# 2. Main Function

```{r, eval=FALSE}
IBaySVG(spelist,c_alpha = NULL,num_cores = 1,acrate = 0.01,alpha = 0.05,
        gamma1 = sqrt(0.001),gamma2 = 0.01,min_iter = 30,max_iter = 200,
        speci_iter = 12,cp = 0.2,dp = 1.8,ak_domain = c(0.08,0.05,0.04,0.03),
        cq = 1,dq = 1,a_phi_single = 0.001,b_phi_single = 0.001,a_pi = 1,
        b_pi = 1,eta_sigma_single = 1,psi_sigma_single = 1)
```


# 3. Input Arguments

| Parameter | Description |
|--------|-----------------|
| `spelist` | A list of length m, where m denotes the number of samples to be integrated. Each element is a sublist containing: (1) expression_matrix (spots × genes), and (2) location_matrix (spots × coordinates). |
| `c_alpha` | A list of m covariate matrices. Default is `NULL` when covariates are not available. |
| `num_cores` | Number of CPU cores used for parallel computation. Default is 1. |
| `acrate` | Minimum threshold for EBLO convergence. Default is 0.01. |
| `alpha` | Significance level for SV gene identification. Default is 0.05. |
| `gamma1` | Hyperparameter of the slab-and-spike prior for regression coefficient $\beta$. Default is `sqrt(0.001)`. |
| `gamma2` |Hyperparameter for the bi-level structure of $\alpha$. To accommodate integration across different numbers of slices, we set $\Gamma_2$ as follows: for more than four slices, $\Gamma_2$ = 0.01; for three slices, $\Gamma_2$ = 0.005; and for two or fewer slices, $\Gamma_2$ = 0.0001. For cross-donor datasets, additional shrinkage is recommended to account for biological heterogeneity: for two donors, 80% of the baseline $\Gamma_2$ value; for three donors, 60%; and for four or more donors, 50% of the baseline. This adjustment helps balance the contribution of individual slices to the posterior probabilities in the iterative hierarchical model.|
| `min_iter` | Minimum number of algorithm iterations. Default is 30. |
| `max_iter` | Maximum number of algorithm iterations. Default is 200. |
| `speci_iter` | Number of times the early convergence criterion must be satisfied before stopping. Default is 12. |
| `cp` |Hyperparameter of alpha's beta prior. Default is 0.2.  |
| `dp` |Hyperparameter of alpha's beta prior. Default is 1.8.  |
| `cq` |Hyperparameter of u's beta prior. Default is 1.  |
| `dq` |Hyperparameter of u's beta prior. Default is 1.  |
| `ak_domain` |Hyperparameter of varience of beta to control the slab and spike prior for different degree. Default are 0.08,0.05,0.04 and 0.03 for degree 1-4, respectively. |
| `a_phi_single` |Hyperparameter of the shape parameter $\phi$. Default is 0.001.   |
| `b_phi_single` |Hyperparameter of the shape parameter $\phi$. Default is 0.001.   |
| `a_pi` |Hyperparameter of the zero-inflation parameter $\pi$. Default is 1.  |
| `b_pi` |Hyperparameter of the zero-inflation parameter $\pi$. Default is 1.  |
| `eta_sigma_single` |Hyperparameter of variance of $\eta$, the baseline gene expression. Default is 1.  |
| `psi_sigma_single` |Hyperparameter of variance of $\psi$, controlling the contribution of covariates. Default is 1.   |


# 4. Output

The function returns a list containing the following components:

| Output | Description |
|--------|------------|
| `initial_result` | Posterior probability $u_k$ in 2D spatial coordinates. |
| `post_mean_uk` | Posterior probability $u_k$ after maximization. |
| `identify_number` | Binary vector (1 = SV gene, 0 = non-SV gene). |
| `svgenename` | Names of identified spatially variable genes. |
|`all_parameters`|Posterior mean of remained parameters.|

### Structure of `all_parameters`

`all_parameters` is a list of length $G$ (number of genes). Each element corresponds to one gene and contains the following posterior summaries:

- **`u`**: Posterior mean of $u_k$ (binary vector).

- **`g`**: A list of length $M$ (number of slices).  
  The $m$-th element contains $g_i^{(m)}$, for $i = 1, \dots, n^{(m)}$.

- **`r`**: A list of length $M$.  
  The $m$-th element contains $r_i^{(m)}$, for $i = 1, \dots, n^{(m)}$.

- **`phi`**: A vector of length $M$, containing $\phi^{(m)}$ for each slice.

- **`sigma_beta`**: An $M \times 2$ matrix.  
  Rows correspond to slices, columns correspond to $k = 1,2$, representing $\sigma_k^{2(m)}$.

- **`a_beta`**: An $M \times 2$ matrix.  
  Rows correspond to slices, columns correspond to $k = 1,2$, representing $a_k^{(m)}$.

- **`alpha`**: An $M \times 2$ matrix.  
  Rows correspond to slices, columns correspond to $k = 1,2$, representing sample-specific indicators $\alpha_k^{(m)}$.

- **`q`**: Binary vector representing $q_k$.

- **`p`**: Binary vector representing $p_k$.

- **`theta`**: A $(1 + 2L + J) \times M$ matrix.  
  Each column corresponds to one slice and contains posterior means of $(\eta^{(m)}, \beta_1^{(m)}, \beta_2^{(m)}, \psi^{(m)})$.

- **`sigma_theta`**: A list of length $M$.  
  Each element is a $(1 + 2L + J) \times (1 + 2L + J)$ covariance matrix, representing the posterior covariance of $\theta^{(m)}$.

# 5. Example Usage

## 5.1  Required R packages:

```{r, eval=FALSE}
truncnorm
nlme
spatstat.data
spatstat.geom
spatstat.random
spatstat.explore
spatstat.model
spatstat.linnet
spatstat
splines
mvtnorm
Rcpp
statmod
R.utils
MCMCpack
Matrix
MASS
parallel
pscl
```

## 5.2  Data Structure Requirements:

Each element of `spelist` must contain:

- `expression_matrix`: matrix of dimension (spots × genes)
- `location_matrix`: matrix of dimension (spots × spatial coordinates)

Example structure:

```{r}
load(here::here("Demo/example","data_example.RData"))
data_example[[1]][[1]][c(1:5),c(1:5)]#expression_matrix where the first 10 genes are set to SV genes and the remained are non-Sv genes
data_example[[1]][[2]][c(1:5),]#location_matrix
```

Each element of `c_alpha` must contain:

- `covariate_matrix`: matrix of dimension (spots × covariates)

Example structure:

```{r}
data_example[[1]][[3]][c(1:5),]#covariate_matrix
```

## 5.3 Integrating the example data with IBaySVG

```{r, message=FALSE, warning=FALSE}
library(truncnorm)
library(nlme)
library(spatstat.data)
library(spatstat.geom)
library(spatstat.random)
library(spatstat.explore)
library(spatstat.model)
library(spatstat.linnet)
library(spatstat)
library(splines)
library(mvtnorm)
library(Rcpp)
library(statmod)
library(R.utils)
library(MCMCpack)
library(Matrix)
library(MASS)
library(parallel)
library(pscl)
source(here::here("Demo","IBaySVG_main.R"))
#or you can download the package from https://github.com/zhoumeng123456/IBaySVG

#expression matrix and location
spelist <- list(list(data_example[[1]][[1]], data_example[[1]][[2]]),
               list(data_example[[2]][[1]], data_example[[2]][[2]]),
               list(data_example[[3]][[1]], data_example[[3]][[2]]),
                list(data_example[[4]][[1]], data_example[[4]][[2]])) 
# covariates
c_alpha <- list(data_example[[1]][[3]],data_example[[2]][[3]],
                data_example[[3]][[3]],data_example[[4]][[3]])
#you can choose c_alpha <- NULL for the lack of cell-type covariates.
result <- IBaySVG(spelist = spelist,c_alpha = c_alpha,num_cores = 10)

# Print key outputs:
# 1. Posterior mean probabilities after maximization
print("Posterior mean probabilities:")
print(result[[2]])

# 2. the identified SV genes
print("the identified SV genes:")
print(result[[4]])

# 3. the posterior mean of parameters
names(result[[5]][[1]])
print("the posterior mean of parameters theta:")
result[[5]][[1]][["theta"]]
```

# 6. Convergence Criteria

The CAVI algorithm adopts three complementary stopping criteria to ensure
numerical stability while avoiding unnecessary computation.

## 6.1 Standard ELBO-Based Convergence

The primary convergence criterion is based on the Evidence Lower Bound (ELBO).
Let $\text{ELBO}^{(t)}$ denote the ELBO at iteration $t$. \textbf{Convergence is declared} if

$$
\left| \text{ELBO}^{(t)} - \text{ELBO}^{(t-1)} \right| < \alpha,
$$

where $\alpha = 0.01$, and the number of iterations exceeds the minimum iteration
threshold (`min_iter = 30`). 

This constraint prevents premature convergence caused by unstable updates during
early iterations.

## 6.2 Early Convergence Based on Posterior Stability

If the absolute change in $u_k$ satisfies $|u_k^{(t)} - u_k^{(t-1)}| < 10^{-9}$ for more than `speci_iter = 12` consecutive iterations, and simultaneously $\left| \text{ELBO}^{(t)} - \text{ELBO}^{(t-1)} \right| < 1,$ the algorithm is \textbf{terminated early}.

This strategy exploits the property that $u_k$ tends to approach 0 or 1, thereby reducing unnecessary computation once posterior probabilities have stabilized.

## 6.3 Maximum Iteration Termination

To prevent potential stagnation in local regions of the parameter space,
the algorithm is \textbf{forcibly terminated} when the number of iterations exceeds
`max_iter = 200`. This safeguard ensures computational efficiency and avoids excessive runtime
in rare pathological cases. In practice, convergence is typically achieved well before reaching the
maximum iteration limit.


# 7. Reproducibility Notes

- All analyses should be executed from the project root directory.
- Avoid modifying the global workspace.
- For multi-core execution, ensure the operating system supports parallel backends.